Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2016-12-17 Master Thesis/synopsis.tex
Add proof for 5.3

\begin{lemma}\label{5.3}
Let $F$ be a number field. Then:
\begin{enumerate}[(i)]
\item $F^\times$ is free-by-finite.
\item Let $F_1,\ldots,F_n$ be finite extensions of $F$. Then $\prod_{i=1}^{n}F_i^\times/F^\times$ is free-by-finite, where $F^\times$ is considered as a subgroup of $\prod_{i=1}^{n}F_i^\times$ via the diagonal embedding.
\end{enumerate}
\end{lemma}

>>>>

\begin{lemma}\label{5.3}
Let $F$ be a number field. Then:
\begin{enumerate}[(i)]
\item $F^\times$ is free-by-finite.
\item Let $F_1,\ldots,F_n$ be finite extensions of $F$. Then $\big(\prod_{i=1}^{n}F_i^\times\big)/F^\times$ is free-by-finite, where $F^\times$ is considered as a subgroup of $\prod_{i=1}^{n}F_i^\times$ via the diagonal embedding.
\end{enumerate}
\end{lemma}

\begin{proof}
\begin{enumerate}[(i)]
\item Let $\mathcal{P}(F)$ be the group of principal ideals of $F$. Consider the following exact sequence:
\[ \begin{tikzcd}
1 \ar[r] & \mathcal{O}_F^\times \ar[r] & F^\times \ar[r]& \mathcal{P}(F) \ar[r] & 1
\end{tikzcd} \]
By Dirichlet's Unit Theorem, $\mathcal{O}_F^\times$ is free-by-finite. $\mathcal{P}(F)$, as a subset of the ideal group of $F$, is free abelian, in particular free-by-finite. The assertion follows from Lemma \ref{5.2} (ii).
\item Consider the following exact sequence:
\[ \begin{tikzcd}
1 \ar[r] & \displaystyle \prod_{i=1}^{n-1} F_i^\times \ar[r]& \displaystyle \Big(\prod_{i=1}^{n}F_i^\times\Big)/F^\times \ar[r]& F_n^\times/F^\times \ar[r]& 1
\end{tikzcd} \]
The left group is free-by-finite by (i), and by Lemma \ref{5.2} (ii), we can use induction to reduce to the case $n=1$.

Let $\mathcal{P}(F)$ denote the group of principal ideals of $F$, $\mathcal{I}(F)$ the ideal group of $F$, and $\operatorname{Cl}(F) = \mathcal{I}(F)/\mathcal{P}(F)$ the ideal class group of $F$. We have the following exact sequence:
\[ \begin{tikzcd}
1 \ar[r]& \mathcal{O}_{F_1}^\times /\mathcal{O}_F^\times\ar[r] & F_1^\times / F^\times\ar[r] & \mathcal{P}(F_1)/\mathcal{P}(F)\ar[r] & 1
\end{tikzcd} \]
Again by Lemma \ref{5.2} (ii), it suffices to show that $\mathcal{P}(F_1)/\mathcal{P}(F)$ is free-by-finite. Using the snake lemma, we get the following exact sequence:
\[
0 \longrightarrow \ker(f)\longrightarrow \mathcal{P}(F_1)/\mathcal{P}(F)\longrightarrow \mathcal{I}(F_1)/\mathcal{I}(F)
\longrightarrow \operatorname{coker}(f)\longrightarrow 0 \]
where $f:\operatorname{Cl}(F)\to \operatorname{Cl}(F_1)$. Since the ideal class groups are finite, according to Lemma \ref{5.2} (i), it suffices to show that $\mathcal{I}(F_1)/\mathcal{I}(F)$ is free-by-finite. This follows from the fact that there are only finitely many primes of $F$ which are ramified in $F_1$.\qedhere
\end{enumerate}
\end{proof}
